---
title: "Big Data Analysis R Assignment"
author: "Kopsco and Ada"
date: "4/3/2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(tidyverse)
library(dbplyr)
library(ggplot2)

download.file("https://ndownloader.figshare.com/files/2292169",
              "data/portal_data_joined.csv")
read_csv("data/portal_data_joined.csv")
read_csv("data/portal_precipitation_19801989.csv")
surveys <- read_csv("data/portal_data_joined.csv")
precip <- read_csv("data/portal_precipitation_19801989.csv")
```

  The Portal, AZ rodent dataset provides information on all rodents sampled among 24 experimental plots in the Chihuahan Desert between 1977 and 2002. The following plot examines the total number of species over the entire observation time. While there is considerable change in total abundance among years, the most abundant species remained relatively stable. *C. baileyi* and *D. merriami* were the most abundant species present, however *C. baileyi* population exploded only in the late 1990s as *D. merriami*'s abundance began to decline.
  
```{r include = FALSE}
total_ann_abund_spp <-surveys %>%
                group_by(year, genus, species) %>%
                tally() 
summary(total_ann_abund_spp)

top_species <- total_ann_abund_spp %>%
              group_by(genus, species, year) %>%
              arrange(desc(n))
```

```{r echo=FALSE}
ggplot(data = total_ann_abund_spp, aes(x = year, y = n, color = species)) +
     geom_line() + theme(axis.text.x = element_text(colour = "grey20", size = 9, angle = 90, hjust =      0.5, vjust = 0.5),axis.text.y = element_text(colour = "grey20", size = 10),
     text=element_text(size = 10))+
     labs(x = "Year of observation",
         y = "Number of species")
```

Overall, the average total number of species increased over the study period. The total average rodent population experienced a slow decline starting in the late 1980s, and then a sharp resurgence in the mid 1990s. 

```{r include=FALSE}
mean_yearly_abund_spp <-surveys %>%
                group_by(year, species) %>%
                tally() %>%
                summarize(mean_abundance_yr = mean(n)) 

summary(mean_yearly_abund_spp)
```

```{r echo=FALSE}
ggplot(data = mean_yearly_abund_spp, aes(x = year, y = mean_abundance_yr)) +
     geom_line() + theme(axis.text.x = element_text(colour = "grey20", size = 9, angle = 90, hjust =      0.5, vjust = 0.5),axis.text.y = element_text(colour = "grey20", size = 10),
     text=element_text(size = 14))+
     labs(title = "Average number of species observed each year",
         x = "Year of observation",
         y = "Mean number of species") 
```

Most plots appear to contain fewer than 60 individuals of a particular species each year. However, *C. baileyi* and *D. merriami* exceed this threshold in several plots in various years. In particular, plot 4 had a surge of *D. merriami* in 1990, and *C. baileyi* increased in the late 1990s in plots 3, 6, 13, 18, 19, 20, 21, 22.

```{r include = FALSE}
spabplot <- surveys %>%
  group_by(species, plot_id, year) %>%
  tally()
```

```{r echo=FALSE}
plot <- ggplot(data = spabplot, aes(x = year, y = n, color = species)) +
     geom_line() +
     facet_wrap(~ plot_id)+ theme(axis.text.x = element_text(colour = "grey20", size = 7, angle = 90,
     hjust = 0.5, vjust = 0.5),axis.text.y = element_text(colour = "grey20", size = 10),
     text=element_text(size = 11))+
     labs(title = "Abundance of species observed per experimental plot",
         x = "Year of observation",
         y = "Number of species") 
plot
```

Plot 4 appears to host the largest abundance of species of all plots. 

```{r echo=FALSE}

plot_count_yr <- surveys %>%
                 group_by(year, species, plot_id) %>%
                 tally()

ggplot(data = plot_count_yr, aes(x = year, y = n)) +
    geom_point() +
    facet_wrap(~ plot_id)+
    theme(axis.text.x = element_text(colour = "grey20", size = 9, angle = 90, hjust = 0.5, vjust = 
    0.5),axis.text.y = element_text(colour = "grey20", size = 10),
    text=element_text(size = 12))+
    labs(title = "Total abundance by experimental plot",
         x = "Year of observation",
         y = "Number of species") 
```

A boxplot of the mean abundance per experimental plot reveals a large number of outliers. A Kruskal-Wallis one way test of variance reveals that there is a significant difference among the mean species abundances per plot_id (p = 'r kruskal.test(species ~ plot_id, data = sp_abund)' ).

```{r include = FALSE}

sp_abund <- surveys %>%
            group_by(plot_id, year, species) %>%
            tally()

mean_abund <- sp_abund %>%
              group_by(plot_id, year) %>%
              summarise(total_abundance = sum(n)) %>%
              summarise(mean_abund = mean(total_abundance))
```

```{r echo=FALSE}
ggplot(data = sp_abund, aes(x = plot_id, y = n, group = plot_id)) +
    geom_boxplot(coef = 1.5) + theme(axis.text.x = element_text(colour = "grey20", size = 12, angle =     90, hjust = 0.5, vjust = 0.5), axis.text.y = element_text(colour = "grey20", size = 12),
    text=element_text(size = 14)) +
    geom_jitter(alpha = 0.3, color = "blue") +
    labs(title = "Species abundance by experimental plot",
         x = "Plot Number",
         y = "Number of species")

kruskal.test(species ~ plot_id, data = sp_abund)
  
```

Of the top five most abundant species, *C. baileyi* and *D. merriami* are the most abundant for all experimental plots. However, there is a lot of variance in their sample numbers throughout the years.

```{r include = FALSE}

most_abundant_species <- surveys %>%
    group_by(species, year, plot_id) %>%
    tally() %>%
    arrange(desc(n)) %>%
    filter(n > 50) 
```

```{r echo=FALSE}
ggplot(data = most_abundant_species, aes(x = year, y = n, color = species)) +
    geom_line() +
    facet_wrap(~ plot_id) +
    theme(axis.text.x = element_text(colour = "grey20", size = 6, angle = 90, hjust = 0.5, vjust =  
    0.5), axis.text.y = element_text(colour = "grey20", size = 10), text=element_text(size = 16))+
        labs(title = "Most abundant species by experimental plot",
         x = "Plot Number",
         y = "Number of species") 

ggplot(data = most_abundant_species, aes(x = species, y = n, group = species)) +
    geom_boxplot(coef = 1.5)+
    theme(axis.text.x = element_text(colour = "grey20", size = 12, angle = 90, hjust = 0.5, vjust = 
    0.5), axis.text.y = element_text(colour = "grey20", size = 12), text=element_text(size = 14))+
    geom_jitter(alpha = 0.3, color = "blue") +
    labs(title = "Most abundance species for all experimental plots",
         x = "Species",
         y = "Species Abundance")

```

Despite differences in plot abundance and overall abundance throughout the study period, there does not appear to be a significant relationship between the annual species abundance and the annual precipitation from 1980-1989 ( p = 'r abund_precip'). Other factors need to be investigated to understand the difference in population dynamics occurring within the plot structure. 

```{r include = FALSE}

surveys_1980_1989 <- surveys %>%
  filter(year < 1989) %>%
  filter(year > 1980)

count_spp_yr <- surveys_1980_1989 %>%
                group_by(year, species) %>%
                tally() 

precip_yr <- precip %>%
            group_by(Year, precipitation) %>%
            tally()

precip_survey <- merge(count_spp_yr, precip_yr) 

abund_precip = glm(n ~ precipitation, family="poisson", data=precip_survey)
```

```{r include=FALSE}
summary(abund_precip)
```

```{r echo=FALSE}
plot(abund_precip)
```

Using a Kruskal Wallace test to compare the amount of precipitation per year, there is no significant difference among sampling years (p = 'r kruskal.test(precipitation ~ year, data = precip_survey)'). This suggests that annual precipitation is not behind the difference in rodent population abundance. 

```{r include=FALSE}

summary(precip_yr)
kruskal.test(precipitation ~ year, data = precip_survey)

```


Average weight per species:

```{r}

clean_weight <- surveys %>% filter(!is.na(weight))

all_weights <- clean_weight %>% select(weight, species_id) %>% group_by(species_id)

p1 <- ggplot(all_weights, aes(x=species_id, y=weight, col = species_id)) + geom_jitter(alpha=.1,aes(color=as.factor(species_id))) + geom_boxplot() + labs(title="Mean weights per species",x="Species ID", y = "Weight") + theme_minimal()

p1
```

This box-plot illustrates mean weights for each species in the survey data.

Summary the mean weights per species:
```{r}
mean_weights_per_species <- clean_weight %>% select(weight, species_id) %>% group_by(species_id) %>% summarise(mean_weight = round(mean(weight),digits = 2))

mean_weights_per_species
```

We choose top five most abundant species and compare mean hindfoot length and weight.

```{r}
top_five_species <- surveys %>% select(species_id, hindfoot_length, weight) %>% filter(!is.na(weight)) %>% filter(!is.na(hindfoot_length)) %>% group_by(species_id) %>% tally()

top_five_species <- top_five_species[order(top_five_species$n),]

top_five_species <- top_five_species[c(20:24),]

#RM, DO, PB, PP, DM

top_five_species <- surveys %>% select(species_id, hindfoot_length, weight) %>% filter(!is.na(weight)) %>% filter(!is.na(hindfoot_length)) 

top_five_species <- top_five_species %>% filter(species_id %in% c("RM", "DO", "PB", "PP", "DM"))

top_five_species$species_id <- as.character(top_five_species$species_id)

a1 <- top_five_species %>% group_by(species_id) %>% summarise(mean_hindfoot = round(mean(hindfoot_length),digits = 2))

a2 <- top_five_species %>% group_by(species_id) %>% summarise(mean_weight = round(mean(weight), digits = 2))

a1$mean_weight <- a2$mean_weight  

means_for_top_five <- a1

p2 <- ggplot(data = means_for_top_five, 
    aes(x=mean_weight, y=mean_hindfoot, color=species_id))+ geom_point() + theme_minimal()

p2
```


This plot shows first five most abundant species and their mean weight vs. mean hindfoot lenghts.    

Summary of the relationship between species' mean weight between mean hindfoot length:

```{r}
test1 <- glm(mean_weight ~ mean_hindfoot, data = means_for_top_five)

test1summary <- summary(test1)

test1summary 
```

There is not significant relationship between species' mean weight between mean hindfoot length (p = 0.00335).

